Cost- and Deadline-constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds

### 2012-01-19 paper/paper.tex
extended sec 2 and added sec 5.1

\section{Related Work}
General policy and rule-based approaches to dynamic provisioning (e.g. Amazon
Auto Scaling\footnote{\url{http://aws.amazon.com/autoscaling}} and
RightScale\footnote{\url{http://www.rightscale.com}}).

Policy-based approaches for scientific workloads (e.g. \cite{Marshall2010, Kim2011}). Our approach is different in that we consider workflows, while policy based approaches typically consider bags of independent tasks or unpredictable batch workloads. This enables us to take advantage of scheduling heuristics that cannot be applied to independent tasks.

Deadline-constrained cost-minimization workflow scheduling. Our work is different from \cite{Yu2005, Abrishami2010} in that we consider ensembles of workflows in IaaS clouds, which allow one to provision resources on a per-hour billing model, rather than utility grids, which allow one to choose from a pool of existing resources with a per-job billing model. Our work is different from \cite{Mao2011} in that we consider ensembles of workflows rather than unpredictable workloads containing workflows. We also have budget constraints rather than cost minimization as a goal. In other words, we assume that there is more work to be done than the available budget, so some work must be rejected. Therefore cost is not something we optimize, but rather a constraint.

Budget-constrained workflow scheduling \cite{Sakellariou2007}.

Bi-criteria scheduling and multi-criteria scheduling. These approaches are similar to ours in that we have two scheduling criteria: cost and makespan. The challenge in multi-criteria scheduling is to derive an objective function that takes into account all of the criteria. These approaches typically use metaheuristics that run for a long time before producing good results.

@@

In this section we describe several algorithms we developed to execute ensembles of
workflows on the cloud under budget and deadline constraints.

\subsection{Static Provisioning Dynamic Scheduling \\
(SPDS) }
\label{sec:spds}

The simplest strategy for executing ensembles of workflows in the cloud is to use
provision resources statically and schedule workflow tasks dynamically. Given a 
budget $b$, deadline $d$, and the hourly price $p$ of a VM it is easy to calculate
the number of VMs, $N_{VM}$, to provision so that the entire budget is consumed before
the deadline:

\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil b / d / p \rceil
\end{equation}

The SPDS algorithm statically provisions VMs at the start of the ensemble execution 
and keeps them running until the deadline is reached, or the budget runs out.
This provisioning plan has the advantage that it minimizes the number of provisioning 
and de-provisioning requests.

Once the VMs are provisioned, the tasks are mapped onto idle VMs using the dynamic
priority-based scheduling procedure shown in Algorithm~\ref{alg:ds}.

\begin{algorithm}
\caption{Priority-based scheduling algorithm for SPDS}
\label{alg:ds}
\begin{algorithmic}[1]
\Procedure{Schedule}{}
    \State $P\gets$ empty priority queue
	\State $IdleVMs\gets$ set of idle VMs
	\For{root task $t$ in all workflows}
    	\State \Call{Insert}{$t,P$} 
    \EndFor
    \While{deadline not reached}
    	\While{$IdleVMs \neq \emptyset\ and\ P \neq \emptyset$}
    		\State $v\gets$ \Call{SelectRandom}{$IdleVMs$}
    		\State $t\gets$ \Call{Pop}{$P$}
    		\State \Call{Submit}{$t,v$}
    	\EndWhile
    	\State Wait for task $t$ to finish on VM $v$
    	\State Update $P$ with ready children of $t$
		\State \Call{Insert}{$v,IdleVMs$}
    \EndWhile
\EndProcedure
\end{algorithmic} 
\end{algorithm}

Initially, the ready tasks from all workflows in the ensemble are added to a 
priority queue based on the priority of the workflow to which they belong. If
there are idle VMs available, and the priority queue is not empty, the next task
from the priority queue is submitted to a randomly chosen idle VM. The process is 
repeated until there are no idle VMs or the priority queue is empty. The scheduler
then waits for a task to finish, adds its ready children to the priority queue,
marks the VM as idle, and the entire process repeats until the deadline is reached.

This algorithm guarantees that tasks from lower priority workflows are
always deferred when higher-priority tasks are available, but lower-priority
tasks can still occupy idle VMs when higher-priority tasks are not available. 
However, because there is no preemption, long-running low-priority tasks may delay 
the execution of higher-priority tasks. In addition, tasks from low priority 
workflows may be executed even though there is no chance that those workflows 
will be completed within the current budget and deadline. Fig.~\ref{fig:spds-example} 
shows an example schedule generated using the SPDS algorithm. The figure illustrates 
how tasks from lower priority workflows backfill idle VMs when tasks from higher 
priority workflows are not available.

\begin{figure}[htb] 
\centering
\includegraphics[width=1.0\columnwidth]{figures/spds-gantt}
 \caption{Example schedule generated using SPDS algorithm. Tasks are colored by workflow. }
\label{fig:spds-example}
\end{figure}

@@

The static provisioning approach of SPDS may not be optimal when resource demand
changes during ensemble execution. In those cases, SPDS may either over-provision, 
wasting the budget on idle VMs, or under-provision, using up the deadline while starving 
the application of usable VMs. One of the advantages of elasticity in cloud computing 
is that the number of resources can be changed to meet the demands of an application.
In policy-based provisioning (or auto-scaling) systems, metrics such as resource 
utilization and queue length are monitored to estimate application demand. The DPDS 
algorithm uses this approach to determine when to provision and de-provision VMs. The 
algorithm periodically computes resource utilization as the percentage of idle VMs, and
adjusts the number of VMs if the utilization is above or below given thresholds.
Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are
approaching their hourly billing cycle when deciding which VMs to terminate. This process
is described in Algorithm~\ref{alg:prov}.

\begin{algorithm}
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
	\State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
    	\State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
    	\State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
    	\State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
		\State $u\gets$ current VM utilization
    	\If{$u>u_h$ and $|V_R| < v_{max}$}
    		\State \Call{Start}{$new\ VM$}
    	\ElsIf{$u<u_l$}
    		\State $V_I\gets$ set of idle VMs
    		\State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
			\State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
    		\State \Call{Terminate}{$V_T$} \label{l:terminate2}
    	\EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

The set of VMs completing their billing cycle is determined by considering both the
provisioner interval, and the termination delay of the provider. This guarantees 
that VMs can be terminated before they start the next billing cycle and prevents the
budget from being overrun. The VMs terminated in line \ref{l:terminate1} are the ones 
that would overrun the budget if not terminated in the current provisioning cycle. 
The VMs terminated on line \ref{l:nT2} are chosen to increase the resource utilization
to the desired threshold. In order to prevent instances from being
terminated too quickly, potentially wasting resources that have already been paid for
but could be used later, no more than half of the idle resources are terminated during
each provisioning cycle. To avoid an uncontrolled increase in the number of
instances, which may happen in the case of highly parallel workflows, the provisioner 
will not start a new VM if the user-defined scaling parameter, $v_{max}$, has
been reached.

@@

The DPDS algorithm does not use any information about the structure of the workflows
in the ensemble when scheduling tasks: it looks only at the priorities of the ready 
tasks when deciding which task to schedule next. It does not consider whether a lower 
priority task belongs to a workflow that it will never be able to complete given the 
current budget and deadline. As a result, DPDS may execute lower priority tasks just 
to keep VMs busy that will end up delaying higher priority tasks, making it less likely
that higher priority workflows will be able to finish.

In order to address this issue, the Workflow-Aware DPDS (WA-DPDS) algorithm extends 
DPDS by introducing a workflow admission procedure. The admission procedure is
invoked whenever WA-DPDS sees the first task of a new workflow at the head of the 
priority queue (i.e. no other tasks from that workflow have been scheduled yet). 
The procedure estimates if there is enough
budget remaining to admit the new workflow: if not, all its tasks are rejected.
Algorithm~\ref{alg:wa-dpds} estimates the cost of admitting a new workflow and
compares it to the current cost (consumed budget) and remaining budget, taking
into account that started VMs have been already paid for. It adds a small safety
margin choosen to avoid overestimation of a budget.

@@

>>>>

\section{Related Work}
General policy and rule-based approaches to dynamic provisioning (e.g. Amazon
Auto Scaling\footnote{\url{http://aws.amazon.com/autoscaling}} and
RightScale\footnote{\url{http://www.rightscale.com}}) allow adjusting the size
of resource pool based on metrics related to infrastructure and application.
Typical infrastructure-specific metric is system load, whereas
application-specific metrics include response time and queue length. It is
possible to set thresholds and limits to tune the behavior of these autoscaling
systems, but no support for complex applications is provided.

Policy-based approaches for scientific workloads (e.g. \cite{Marshall2010,
Kim2011}) also allow to scale the the cloud resource pool or to extend the
capabilities of clusters using cloud-burst techniqies. Our approach is different
in that we consider workflows, while policy based approaches typically consider
bags of independent tasks or unpredictable batch workloads. This enables us to
take advantage of scheduling heuristics that cannot be applied to independent
tasks.

Our work is related to the strategies for deadline-constrained cost-minimization
workflow scheduling, developed for utility grid systems. However, our problem is
different from \cite{Yu2005, Abrishami2010} in that we consider ensembles of workflows in
IaaS clouds, which allow one to provision resources on a per-hour billing model,
rather than utility grids, which allow one to choose from a pool of existing
resources with a per-job billing model. Our work is also different from
cloud-targeted autoscaling solution~\cite{Mao2011} in that we consider ensembles
of workflows rather than unpredictable workloads containing workflows. We also have budget constraints
rather than cost minimization as a goal. In other words, we assume that there is
more work to be done than the available budget, so some work must be rejected.
Therefore cost is not something we optimize, but rather a constraint.

We should refer to bi-criteria scheduling and multi-criteria scheduling of
workflows~\cite{Wieczorek2009,Prodan10,Dongarra2007}. These approaches are
similar to ours in that we have two scheduling criteria: cost and makespan. The
challenge in multi-criteria scheduling is to derive an objective function that
takes into account all of the criteria. In our case one objective (amount
of work completed) is subject to optimization, whereas time and cost are
trated as constraints. Other approaches~\cite{Talukder2009,Pandey2010} use
metaheuristics that usually run for a long time before producing good results,
which makes tham less useful in practical applications. Our work can be also
regarded as extension of the budget-constrained workflow scheduling
\cite{Sakellariou2007} in the sense that we are dealing with workflow ensembles
and the deadline constraint is added.

@@

\subsection{Static Provisioning Dynamic Scheduling\\
(SPDS) }
\label{sec:spds}
This is the simplest strategy based on static provisioning of resources and
on-line scheduling of workflow tasks. Given the budget $B$ and deadline $d$
together with the hourly price $p$ of a VM it is possible to compute the number
of VMs $N_{VM}$ to provision so that they consume the whole budget before the
deadline.

\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil B / d / p \rceil
\end{equation}

Although simple, such provisioning plan has the advantage that it
minimizes the number of provisioning and deprovisioning requests.
Once the resources are provisioned, the tasks are scheduled to VMs using dynamic
priority-based Algorithm~\ref{alg:ds}.

\begin{algorithm}
\caption{Priority-based scheduling}
\label{alg:ds}
\begin{algorithmic}[1]
\Procedure{schedule}{}
    \State $Q\gets set\ of\ released\ tasks$
    \State $P\gets empty\ priority\ queue$
    \While{$Q \neq \emptyset$}
    	\For{$task\ t\in Q$}
    		\State \Call{Insert}{$t,P$} 
    	\EndFor
    	\While{$FreeVMs \neq \emptyset \wedge P \neq \emptyset $}
    		\State $VM\gets$ \Call{SelectRandom}{$FreeVMs$}
    		\State $t\gets$ \Call{Pop}{$P$}
    		\State \Call{Submit}{$t,VM$}
    	\EndWhile
    	\State \Call{Wait}{$task\ finished$}
    	\State \Call{Update}{$Q$}
    \EndWhile
\EndProcedure
\end{algorithmic} 
\end{algorithm}

Initially all the
released tasks from all the workflows are submitted to the queue $Q$. Then, they
are inserted into priority queue $P$ based on the priority of workflows. If
there are free VMs available, the tasks from the priority queue are submitted to
the randomly choosen VMs. The process is repeated each time a task finishes.
Using the priority queue guarantees that tasks from lower priority workflows are
always submitted later than the higher priority ones and that the lower priority
tasks can occupy idle VMs. However, as there is no preemption, long-running
low-priority tasks may delay execution of high-priority tasks when such ones
become ready. Fig.~\ref{fig:spds-example} shows the example plan generated using
SPDS algorithm. It can be seen that tasks from lower priority workflows can fill
idle time of provisioned virtual machines.

\begin{figure}[htb] 
\centering
\includegraphics[width=1.0\columnwidth]{figures/spds-gantt}
 \caption{Example of schedule generated using SPDS algorithm: tasks are labelled
 by colors depending on the workflow. }
\label{fig:spds-example}
\end{figure}

@@

Static provisioning plan used in SPDS algorithm allows no changes to resource
pool at runtime. However, such plan may not be optimal when resource demand
changes during execution of workflows. In policy-based provisioning
(or auto-scaling) systems the metrics such as resource
utilization or queue length are typicaly used to estimate curret resource
demand. Our DPDS algorithm is based on resource utilization: the provisioner
module periodically computes utilization as the percentage of idle VMs. The
provisioner algorithm detects which VMs are finishing their hourly billing cycle
during the next provisioning period and decides whether they should continue
running, should be terminated or new VMs should be provisioned: see
Algorithm~\ref{alg:prov}.

\begin{algorithm}
\caption{Dynamic provisioner}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization thershold; $u_l$: lower utilization
threshold.
\Procedure{provision}{}
    \State $V_C\gets set\ of\ VMs\ completing\ billing\ cycle$
    \State $n_T\gets\ 0$ \Comment{number of VMs to terminate} 
    \If{$ b - c < p * |V_C| \vee t > d $ }
    	\State $V_R\gets set\ of\ running\ VMs$
    	\State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$ 
    	\State $V_T\gets select\ n_T VMs\ to\ terminate\ from\ V_C$
    	\State \Call{Terminate}{$V_T$}
    \Else 
    	\If{$u>u_h$}
    		\State \Call{Start}{$new\ VM$}
    	\EndIf
    	\If{$u<u_l$}
    		\State $V_F\gets free\ (idle)\ VMs$
    		\State $n_T\gets \lceil|V_F|/2\rceil$ 
    		\State \Call{Terminate}{$V_T$}
    	\EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

The number of instances to terminate $n_T$ computed in line 6 is the number that
would overrun the budget if not terminated. On the other hand $n_T$ computed in
line 15 prevents terminating all the instances too quickly, but makes sure that
if there is only one instance left it should be terminated. However, the
termination process is not immediate: to make sure that VMs that have been
already paid for are not terminated prematurely the {\em Terminate()} procedure
terminates only these instances that are close to the end of their billing
cycle. The remaining time threshold is calculated as a sum of provisioner
interval and estimated termination delay. This guarantees that no VMs can
overrun the assigned budget. To avoid uncontrolled increase of number of
instances, which may happen in the case of highly parallel workflows,
provisioner checks also the maximum scaling factor from the initial number of
provisioned VMs and prohibits starting new ones when this threshold is reached.

@@

DPDS algorithm does not use any information about workflow structure: it deals
only with scheduling of individual released/ready tasks using their
workflow priorities. If there are no high priority tasks in the ready queue,
then lower priority tasks are submitted to idle resources. This may lead to the
deficiency in the following scenario: if a long-running task of low
priority is submitted to the idle machine and then several high
priority tasks are released to the ready queue, they become delayed by the
execution of the low priority task. This usually does not cause a problem when
the number of tasks and resources is high, but when the execution approaches the
budget and deadline constraint, we have observed that tasks of low-priority
workflows may prevent higher priority workflows from completion. This
observation leads to the improvemed algorithm which we call workflow-aware DPDS
(WA-DPDS).

WA-DPDS extends DPDS by introducing a workflow admission procedure, which is
invoked each time a task of a new workflow is at the head of priority queue,
i.e. just before it is submitted. The procedure estimates if there is enough
budget remaining to admit a new workflow: if not, all its tasks are rejected.
Algorithm~\ref{alg:wa-dpds} estimates the cost of admitting a new workflow and
compares it to the current cost (consumed budget) and remaining budget, taking
into account that started VMs have been already paid for. It adds a small safety
margin choosen to avoid overestimation of a budget.

@@

To evaluate our algorithms, we developed a cloud workflows simulator based on
CloudSim~\cite{Calheiros11}. To represent our infrastructure model, we defined
Cloud, VM and WorkflowEngine simulation entities. Cloud object is responsible
for staring and terminating VMs using API similar to Amazon EC2. VM object is
responsible for execution of individual tasks, while WorkflowEngine deals with
management of ensembles and workflows. WorkflowEngine uses separate, although
tightly coupled Scheduler and Provisioner modules. We assume that a single core
of a VM executed submitted tasks sequentially and that a task is given exclusive
access to allocated VM. Although CloudSim provides a more advanced
infrastructure model which includes time-sharing and space-sharing policies and
power consumption in a datacenter, we do not use these features, since we are
interested mainly in execution of tasks on VMs and high-level worklow scheduling
on top of that. The simulator relies on DAG and DAX format used by Pegasus
project and workflow gallery.
